\documentclass[12pt,a4paper]{article}

% ============================================================
% PACKAGES
% ============================================================
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\usepackage{colortbl}
\usepackage{xcolor}
\usepackage{makecell}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{amsmath}
\usepackage{tcolorbox}

\definecolor{headerblue}{RGB}{213,220,228}

\newcolumntype{Y}{>{\centering\arraybackslash}X}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}

\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt}

\pagestyle{fancy}
\fancyhf{}
\rhead{SE: Machine Learning Lab}
\lhead{Lab 13}
\rfoot{\thepage}

\begin{document}

% ============================================================
% HEADER
% ============================================================
\renewcommand{\arraystretch}{2.0}
\begin{tabularx}{\textwidth}{|Y|}
\hline
\rowcolor{headerblue} \textbf{\large Department of Computer and Software Engineering} \\
\hline
\textbf{\large SE: Machine Learning} \\
\hline
\end{tabularx}
\renewcommand{\arraystretch}{1.0}

\vspace{12pt}

% ============================================================
% COURSE INFO
% ============================================================
\renewcommand{\arraystretch}{2.0}
\begin{tabularx}{\textwidth}{|L|L|}
\hline
\textbf{Course Instructor:} & \textbf{Date:} \\
\hline
\textbf{Day:} & \textbf{Semester:} \\
\hline
\end{tabularx}
\renewcommand{\arraystretch}{1.0}

\vspace{12pt}

% ============================================================
% TITLE
% ============================================================
\begin{center}
{\Large\bfseries Lab 13: Drift Detection in Streaming Data}
\end{center}

\vspace{6pt}

% ============================================================
% META TABLE
% ============================================================
\renewcommand{\arraystretch}{1.8}
\begin{tabularx}{\textwidth}{|Y|c|c|c|c|}
\hline
\textbf{Lab Title} & \textbf{CLO} & \textbf{BT} & \textbf{PLO} & \textbf{Weightage} \\
\hline
\makecell{{\small Drift Detection in Streaming Data} \\ {\small KS Test, Alerts \& Concept Drift}} & & & & \\
\hline
\end{tabularx}
\renewcommand{\arraystretch}{1.0}

\vspace{6pt}

% ============================================================
% STUDENT TABLE
% ============================================================
\renewcommand{\arraystretch}{2.0}
\begin{tabularx}{\textwidth}{|Y|Y|Y|Y|Y|}
\hline
\textbf{Name} & \textbf{Roll Number} & \textbf{Report /10} & \textbf{Viva /5} & \textbf{Total /15} \\
\hline
 & & & & \\
\hline
\end{tabularx}
\renewcommand{\arraystretch}{1.0}

\vspace{12pt}

\begin{flushright}
Checked on: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

\vspace{6pt}
Signature: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
\end{flushright}

\newpage

% ============================================================
% OBJECTIVES
% ============================================================

\section*{Objectives}

\begin{itemize}
\item Load and explore a structured transaction dataset with controlled drift
\item Simulate streaming data by slicing the dataset into monthly batches
\item Detect statistical distribution shifts using threshold-based methods
\item Apply the Kolmogorov-Smirnov (KS) test for formal drift detection
\item Understand feature drift vs.\ concept drift
\item Observe how concept drift degrades a trained classifier over time
\end{itemize}

% ============================================================
% DATASET
% ============================================================

\section*{Dataset}

The dataset for this lab is generated using the provided Python script \texttt{generate\_drift\_dataset.py}. Running the script produces a file named \texttt{transactions\_with\_drift.xlsx} containing synthetic online payment transactions over a six-month period (January--June 2024).

The file contains three sheets:

\renewcommand{\arraystretch}{1.6}
\begin{tabularx}{\textwidth}{|l|L|}
\hline
\textbf{Sheet} & \textbf{Description} \\
\hline
\texttt{All\_Transactions} & 3{,}000 rows across all six months, sorted by timestamp. Use this as your streaming data source. \\
\hline
\texttt{Baseline\_M1\_M3} & Months 1--3 only (1{,}500 rows). Use this as the reference distribution for all drift comparisons. \\
\hline
\texttt{Drift\_Summary\_Ground\_Truth} & Per-month true statistics. Consult this \textbf{only after completing all tasks} to verify your results. \\
\hline
\end{tabularx}
\renewcommand{\arraystretch}{1.0}

\vspace{8pt}

The dataset contains the following columns:

\renewcommand{\arraystretch}{1.6}
\begin{tabularx}{\textwidth}{|l|L|}
\hline
\textbf{Column} & \textbf{Description} \\
\hline
\texttt{transaction\_id} & Unique transaction identifier \\
\hline
\texttt{timestamp} & Date and time of the transaction \\
\hline
\texttt{month} & Month number (1--6); use this to split data into batches \\
\hline
\texttt{transaction\_amount} & Transaction amount in USD; primary drift feature \\
\hline
\texttt{customer\_age} & Age of the customer in years \\
\hline
\texttt{transaction\_hour} & Hour of day (0--23); shifts toward later hours under drift \\
\hline
\texttt{device\_risk\_score} & Device risk score (0.0--1.0); increases under drift \\
\hline
\texttt{is\_fraud} & Binary fraud label: 1 = fraud, 0 = legitimate \\
\hline
\texttt{drift\_phase} & Phase label: \textit{Baseline}, \textit{Feature Drift}, or \textit{Concept Drift} \\
\hline
\end{tabularx}
\renewcommand{\arraystretch}{1.0}

% ============================================================
% PROBLEM STATEMENT
% ============================================================

\section*{Problem Statement}

Modern machine learning systems operate in continuously changing environments where incoming data may gradually differ from the data used during training.

Examples include:
\begin{itemize}
\item Fraud patterns changing over time
\item Customer behaviour evolving seasonally
\item Sensor calibration drift in IoT systems
\item Changing traffic conditions in ride-sharing systems
\end{itemize}

In this lab, you are a data scientist at an online payment platform. The fraud detection model was trained on transaction data from January to March 2024 (Months 1--3). From April onwards, customer spending behaviour began to shift --- transaction amounts increased, customers transacted at later hours, device risk scores rose, and a new wave of fraud emerged with characteristics different from the training period.

Your task is to build a drift detection pipeline that:
\begin{itemize}
\item Treats each month as an incoming streaming batch
\item Detects distribution shifts in transaction features
\item Raises alerts when drift is detected
\item Logs drifting batches with supporting statistics
\item Uses the KS test for statistical comparison
\item Observes concept drift effects on model performance
\end{itemize}

You will work with a transaction dataset containing:
\begin{itemize}
\item Transaction amount
\item Customer age
\item Transaction hour
\item Device risk score
\item Fraud label
\end{itemize}

% ============================================================
% TASKS
% ============================================================

\section*{Part A: Streaming Data Simulation}

\begin{tcolorbox}
\textbf{Task 1: Load Baseline Dataset}

Load the \texttt{Baseline\_M1\_M3} sheet and compute the following for each numerical feature (\texttt{transaction\_amount}, \texttt{customer\_age}, \texttt{transaction\_hour}, \texttt{device\_risk\_score}):
\begin{itemize}
\item Mean
\item Standard deviation
\item Minimum and maximum
\end{itemize}

Also compute the overall fraud rate (proportion of rows where \texttt{is\_fraud} = 1).

Record these values carefully --- they serve as your reference statistics for all drift detection tasks.

\vspace{2cm}
\vspace{2cm}
\end{tcolorbox}

\vspace{6pt}

\begin{tcolorbox}
\textbf{Task 2: Simulate Streaming Batches}

Load the \texttt{All\_Transactions} sheet and split the data into six monthly batches using the \texttt{month} column. Each batch represents one month of incoming transactions.

For each batch, compute the mean of \texttt{transaction\_amount} and the mean of \texttt{transaction\_hour}. Visualize how these two feature means evolve across the six months using line plots.

\vspace{2cm}
\vspace{2cm}
\end{tcolorbox}

% ============================================================

\section*{Part B: Threshold-Based Drift Detection}

\begin{tcolorbox}
\textbf{Task 3: Mean Shift Detection}

Using the baseline mean and standard deviation of \texttt{transaction\_amount} computed in Task 1, implement a drift detector that compares each monthly batch mean against the baseline.

Raise a drift alert for a batch if:

\[
|\,\mu_{batch} - \mu_{baseline}\,| > 2\,\sigma_{baseline}
\]

where:
\begin{itemize}
\item $\mu_{batch}$ = mean of \texttt{transaction\_amount} in the current monthly batch
\item $\mu_{baseline}$ = mean of \texttt{transaction\_amount} in Months 1--3
\item $\sigma_{baseline}$ = standard deviation of \texttt{transaction\_amount} in Months 1--3
\end{itemize}

Apply this check to all six monthly batches and record which months trigger an alert.

\vspace{2cm}
\vspace{2cm}
\end{tcolorbox}

\vspace{6pt}

\begin{tcolorbox}
\textbf{Task 4: Drift Logging}

For every month that triggers a drift alert in Task 3, store and display the following information:
\begin{itemize}
\item Batch ID (month number)
\item Drift phase (\textit{Baseline}, \textit{Feature Drift}, or \textit{Concept Drift})
\item Batch mean of \texttt{transaction\_amount}
\item Shift magnitude: $|\,\mu_{batch} - \mu_{baseline}\,|$
\item Number of standard deviations from baseline: $\dfrac{|\,\mu_{batch} - \mu_{baseline}\,|}{\sigma_{baseline}}$
\end{itemize}

Display a summary table of all detected drift events. After completing all tasks, verify your results against the \texttt{Drift\_Summary\_Ground\_Truth} sheet.

\vspace{2cm}
\vspace{2cm}
\end{tcolorbox}

% ============================================================

\section*{Part C: KS Test}

\begin{tcolorbox}
\textbf{Task 5: Apply KS Test}

Use the Kolmogorov-Smirnov (KS) test to compare the \texttt{transaction\_amount} distribution of the baseline (Months 1--3) against each of the six monthly batches.

For each batch, report:
\begin{itemize}
\item KS statistic
\item p-value
\end{itemize}

Flag a batch as drifted if the p-value falls below 0.05. Record which months are flagged.

\vspace{2cm}
\vspace{2cm}
\end{tcolorbox}

\vspace{6pt}

\begin{tcolorbox}
\textbf{Task 6: Compare Statistical and Threshold Methods}

Construct a comparison table with one row per month, showing the mean-shift alert (Task 3) and KS test alert (Task 5) side by side.

Discuss:
\begin{itemize}
\item Do both methods agree on which months show drift?
\item Which method detects drift earlier?
\item Sensitivity, false alarm rate, and robustness of each method
\end{itemize}

\vspace{2cm}
\vspace{2cm}
\end{tcolorbox}

% ============================================================

\section*{Part D: Concept Drift}

\begin{tcolorbox}
\textbf{Task 7: Train Baseline Model}

Train a logistics regression classifier on the \texttt{Baseline\_M1\_M3} data to predict \texttt{is\_fraud}.

Use the four numerical features: \texttt{transaction\_amount}, \texttt{customer\_age}, \texttt{transaction\_hour}, and \texttt{device\_risk\_score}.

Evaluate on a held-out 20\% split of the baseline data and report:
\begin{itemize}
\item Accuracy
\item F1-score
\end{itemize}

This model represents the production fraud detector before any drift occurred.

\vspace{2cm}
\vspace{2cm}
\end{tcolorbox}

\vspace{6pt}

\begin{tcolorbox}
\textbf{Task 8: Evaluate on Drifted Data}

Without retraining, evaluate the same model from Task 7 on each of the six monthly batches separately.

For each batch, record:
\begin{itemize}
\item Accuracy
\item F1-score
\end{itemize}

Observe how model performance changes as drift increases across months. Pay particular attention to Month 6, where the fraud rate triples compared to the baseline period.

\vspace{2cm}
\vspace{2cm}
\end{tcolorbox}

% ============================================================

\section*{Part E: Visualization}

\begin{tcolorbox}
\textbf{Task 9: Drift Visualization}

Create a single figure with four subplots summarising your findings across all six months:

\begin{itemize}
\item Mean of \texttt{transaction\_amount} per month, with the $2\sigma$ alert threshold marked
\item KS statistic per month, with the $p = 0.05$ significance boundary marked
\item Fraud rate per month (proportion of \texttt{is\_fraud} = 1)
\item Model F1-score per month from Task 8
\end{itemize}

Shade the background of Months 4--5 to indicate Feature Drift and Month 6 to indicate Concept Drift across all four subplots.

\vspace{2cm}
\vspace{2cm}
\end{tcolorbox}

% ============================================================

\section*{Discussion Questions}

\begin{itemize}
\item What is the difference between feature drift and concept drift? Use specific months from this dataset to illustrate your answer.
\item In Month 4, transaction amounts increased but the fraud rate only rose slightly. Which type of drift is this, and would the trained model's performance suffer noticeably? Why or why not?
\item In Month 6, the fraud rate tripled relative to the baseline. Why does this cause the sharpest drop in F1-score even if accuracy does not fall as dramatically?
\item Why might a simple mean-shift detector miss a drift event that the KS test successfully catches? Give a concrete example.
\item How can a production ML system respond to detected drift automatically? Describe one practical strategy.
\end{itemize}

\end{document}